\documentclass[11pt, a4paper]{article}

\usepackage[
includeheadfoot,
top = 1cm,
bottom = 2cm,
left = 2cm,
right = 2cm,
headsep = 1.5cm,
tmargin = 1cm,
headheight = 0.5cm]{geometry}

\usepackage[style=numeric,
backend=biber, 	%Einstellung des verwendeten Dateityps der Literaturdatenbank
isbn = false, 	%Keine Angabe der ISBN im Literaturverzeichnis
doi = true,  	%Keine Angabe der DOI im Literaturverzeichnis
url = false,	%Angabe der URL im Literturverzeichnis
maxbibnames = 99,
urldate = comp,
sorting = none]{biblatex}

\addbibresource{own_publications.bib}


\usepackage[activate={true,nocompatibility},
final,
tracking=true,
factor=1100,
stretch=10,
shrink=10]{microtype}

\usepackage{url}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[utf8]{inputenc}
\usepackage{paralist}
\usepackage[onehalfspacing]{setspace}
\usepackage[english]{babel}
\usepackage{paralist}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{eurosym}
\usepackage{longtable}
\usepackage{array}

\usepackage{fancyhdr}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}

\usepackage{tikz}
\usetikzlibrary{arrows,positioning} 
\tikzset{
	%Define standard arrow tip
	>=stealth',
	%Define style for boxes
	punkt/.style={
		rectangle,
		rounded corners,
		draw=black, very thick,
		text width=7em,
		minimum height=2em,
		text centered},
	% Define arrow style
	pil/.style={
		->,
		thick,
		shorten <=2pt,
		shorten >=2pt,}
}

\begin{document}
	\fancyhead[C]{\textbf{Joshua Philipp Entrop}\\		
		\footnotesize %Bygatan 37, SE-171 49 Solna · Phone: (+46) 76 255 20 58 \\ 
		Email: joshuaentrop@posteo.de · URL: www.joshua-entrop.com \\}
	
	\begin{center}
		\textbf{Introduction}
	\end{center}
	
	\noindent I am a Ph.D. student with a passion for health data and statistical analyses that have an impact on people's health and well-being. During my bachelor studies in social science, I developed my interest in population-based health research. In my master studies, I focused more specifically on clinical epidemiological research, due to my interest in the application of complex survival models. Currently, I am working as a Ph.D. student and statistician in the field of lymphoma epidemiology.
	
	\begin{center}
		\textbf{Curriculum Vitae}
	\end{center}
	
	\begin{longtable}{p{4cm}p{12cm}}
		
		\textbf{Working experience} &
		\\ \hline
		04/2021 - ongoing 			& \textbf{Statistician}, Karolinska Institutet, Division of Clinical Epidemiology, Department of Medicine Solna \newline
		
		Statistical consulting for internal and external research projects. Database management for the Swedish LymphomaBaSe linkage including checking, cleaning, preparing and distributing data for research projects involving data from LymphomaBaSe. \\
		&							\\
		
		\textbf{Academic Education} & 	\\ \hline
		09/2020 - ongoing			& \textbf{Ph.D. Student in Medical Science}, Karolinska Institutet, Division of Clinical Epidemiology, Department of Medicine Solna \newline
		
		Thesis topic: Development and application of modern statistical methods for studies of childbearing among lymphoma survivors (main {supervisor}: Dr. Sandra Eloranta) \newline 
		
		Focus:
		\begin{compactitem}
			\item Biostatistics, especially survival analysis
			\item Causal inference
			\item Reproducibility in research
			\item Lymphoma research
		\end{compactitem} \\
		
		09/2018 - 06/2020			& \textbf{Public Health Epidemiology (M.Med.Sc.)}, Karolinska Institutet \newline
		Grade: 102.5 CP passed with distinction, 17.5 CP passed
		\newline
		
		Master thesis: The association between sarcoidosis and type 2 diabetes mellitus stratified by corticosteroid treatment at sarcoidosis diagnosis: a registry-based matched cohort study (main supervisor: Dr. Elizabeth Arkema) \newline
																						\\
		04/2015 - 05/2018			& \textbf{Social Science (B.A.)}, Ruhr-University Bochum, Germany \newline
		Grade: 1.2 (Passed with distinction, best graduated student in 2018) \newline
		
		Bachelor thesis: Economic, social, cultural capital and disease. An empiric examination of Cockerham’s Health Lifestyle Theory based on Data of the ALLBUS 2014 (main supervisor: Prof. Dr. Cornelia Weins) \newline                                                									\\ 
				
		\textbf{Scholarships/Awards} 		& 																    \\ \hline
		08/2023 & \textbf{Student Prize for Best Conference Abstract}, International Society of Clinical Biostatistics (ISCB) \\
		09/2021 & \textbf{Price for Best Master Thesis}, German Association for Medical Informatics, Biometry and Epidemiology (GMDS) e. V.\\ 
		09/2019 - 06/2020			& \textbf{Scholarship Holder}, German Academic Exchange Service (DAAD) 		\\
		& Merit-based scholarship                                               								\\
		04/2016 - 06/2020			& \textbf{Scholarship Holder}, Friederich-Ebert-Foundation 					\\
		& Merit-based scholarship                                               								\\													                								
		
		&																										\\ 
		
		\textbf{Software}  		&																			\\ \hline
		 & \textbf{JointFPM}, An R-package implementing flexible parametric model for estimating the mean number of events in the presence of competing risks. Available on CRAN.	\\ 
		 & \textbf{TreeMineR}, An R-package implementing tree-based scan statistics, e.g., for identifying disease clusters along the ICD-10 tree. Available on GitHub. \\ 
		 & \textbf{ExclusionTable}, An R-package that helps creating counting the number of individuals included or excluded from a study population. Available on CRAN. \\ 

		\textbf{IT-Knowledge}		& 																			\\ \hline
		8 years of experience       & MS Office 																\\
		6 years of experience		& R, \LaTeX																	\\					
		4 years of experience		& Stata, Git, GNU Make														\\
		Basic knowledge				& SAS, C++, Python															\\
		&																										\\
		\textbf{Languages}  		&																			\\ \hline
		& German: Native Speaker																				\\
		& English: C1 CEFR																						\\
		& Swedish: B2 CEFR																						\\
		&																										\\
		\textbf{Interests}          &                                                                           \\ \hline
		& Hiking, rowing, equestrian vaulting, cinema, \textit{R}           						\\
	\end{longtable}
	\vspace{0.5cm}
	
	\newpage
	\nocite{*}
	\LARGE{\textbf{Publications}}
	\vspace*{-2cm}
	\printbibliography[title = {\mbox{}}]
	
\end{document}